Bootstrapping Example
========================================================
Bootstrap scripts written by Desmond Ong (dco@stanford), converted to R Markdown by Steph Gagnon

Load in `doBootstrap` functions:
-------------------------------
```{r}
source("http://www.stanford.edu/class/psych252/tutorials/doBootstrap.R")
```

`doBoot` is the basic function that calculates descriptive statistics. For example, you can use the following format:

> results <- doBoot(x, y=NULL, mediator=NULL, whichTest = "mean", numberOfIterations = 5000, confidenceInterval=.95, na.rm=TRUE)

**Output:**

- **Results**

**Input:**

+ **x** = first data vector. Feed in a VECTOR!
+ **y** = second data vector (not necessary if you're doing statistics on one data vector only, e.g. mean)
+ **mediator** = mediator data vector (if you want to do mediation)
+ **whichTest** = the test you want to do, as a string.
    - "mean" (single vector test)
    - "correlation" (correlation between two vectors, paired)
    - "difference, unpaired" (difference between two vectors, unpaired)
    - "difference, paired" (difference between two vectors, paired)
    - "cohen, unpaired" (basically unpaired difference / pooled standard dev) * not bias corrected   nor accelerated
    - "mediation" (calculates a x-->y, x-->med-->y mediation, using Benoit's bm.bootmed code)
    - "custom function": allows you to supply your own function, such that the desired statistic is customFunction(x) or customFunction(x,y) if y is supplied.
+ **numberOfIterations** = number of bootstrapping iterations. Default is 5000
+ **confidenceInterval** = specifies the percentile of the CI. Default is .95
+ **na.rm** = remove NAs?

Some examples with mtcars dataset
------------------------------
First, let's load in the $mtcars$ dataset
```{r}
head(mtcars)
```

### Bootstrapping the Mean
Now, let's try bootstrapping the mean miles per gallon ($mpg$) of the cars in the mtcars dataset.  Note that if you don't provide an argument to $whichTest$ the function won't run!
```{r}
#resultsMean <- doBoot(mtcars$mpg) #error!
resultsMean <- doBoot(mtcars$mpg, whichTest = "mean")
```
Here, we can see that the bootstrapped estimate for mean miles per gallon is `r resultsMean$value`, and the 95% confidence interval of this estimate is from `r resultsMean$ci.low` to `r resultsMean$ci.high`.

If we compare this to the actual data, we can see that the bootstrapped mean and 95% CI closely match that observed in the data.
```{r fig.width=5, fig.height=4}
boxplot(mtcars$mpg, notch=TRUE, col='dodgerblue', ylab='mpg')
```


### Bootstrap a correlation:
```{r}
resultsCorrelation <- doBoot(mtcars$disp,mtcars$hp, whichTest = "correlation")
```
Here we bootstrapped a correlation between cars' displacement (`disp`) and gross horsepower (`hp`). The function returned an estimated correlation of `r resultsCorrelation$value` with a 95% CI from `r resultsCorrelation$ci.low` to `r resultsCorrelation$ci.high`.

Let's observe the correlation using the data in our sample:
```{r fig.width=5, fig.height=5}
plot(mtcars$disp,mtcars$hp, col='firebrick', pch=16)
```


### Bootstap a difference between Means
Next, let's bootstrap the **difference** between the mean displacement and horsepower. Here, we can either treat the values as **unpaired** (i.e., coming from independent sources), or **paired** (i.e., both values come from the same car). First, let's check out the data:
```{r fig.width=5, fig.height=5}
boxplot(mtcars$disp,mtcars$hp, names=c('disp', 'hp'), col=c('mediumseagreen', 'mediumvioletred'))
```

Now let's estimate the difference between the mean displacement and horsepower:
```{r}
resultsDifference_u <- doBoot(mtcars$disp,mtcars$hp, whichTest = "difference, unpaired")
resultsDifference_p <- doBoot(mtcars$disp,mtcars$hp, whichTest = "difference, paired")
```

Here, we can see that both tests calculate a similar difference in means (diff = `r resultsDifference_u`), but different confidence intervals. Assuming the mean displacement and horsepower are **unpaired**, the 95% CI is larger, `r resultsDifference_u$ci.low` - `r resultsDifference_u$ci.high`, whereas when we take into account that our observations of displacement and horsepower come from the same car (i.e., **paired**), the 95% CI is smaller `r resultsDifference_p$ci.low` - `r resultsDifference_p$ci.high`. Neither 95% CI overlaps with zero, and so we might reject the null hypothesis that the difference between the mean displacement and mean horsepower is equal to zero; in other words, it appears that the mean displacement is higher than the mean horsepower. 

### Bootstrap Cohen's D
```{r}
resultsCohen <- doBoot(mtcars$disp,mtcars$hp, whichTest = "cohen, unpaired")
resultsCohen <- doBoot(mtcars$disp,mtcars$hp, whichTest = "cohen, paired")
```

### Bootstrap Mediation using Benoit's code
```{r fig.width = 5, fig.height=4}
resultsMediation <- doBoot(mtcars$disp,mtcars$hp, mtcars$cyl, whichTest = "mediation")
```


Bootstrapping Regressions with `doBootRegression`
---------------------------
`doBootRegression` is the function you call if you want to bootstrap regressions.

> results <- doBootRegression(dataset, formula, mixedEffects = FALSE, numberOfIterations = 5000, confidenceInterval=.95, na.rm=TRUE)

**Output:**
+ **Results**

**Input:**
+ **dataset** = input dataset
+ **formula** = the regression formula of interest. For e.g., "y ~ x" or "y ~ x + (1|z)"
+ **mixedEffects** = whether your formula has mixed effects. Default FALSE.
    - If true, will use lmer, otherwise, lm. 
    - For lmer, will output only fixed effect coefficients
    - For lm, will output coefficients, and R^2
+ **numberOfIterations** = number of bootstrapping iterations. Default is 5000
+ **confidenceInterval** = specifies the percentile of the CI. Default is .95
+ **na.rm** = remove NAs?

```{r}
resultsRegression <- doBootRegression(mtcars, disp~hp)
#resultsRegression <- doBootRegression(mtcars, disp~hp+(1|cyl), mixedEffects = TRUE)
```



