Section, Week 1
=======================================
  
Question 1
-----------------------------------------

### 1.b-i)
**Calculate stats for maltreatment deaths:**
```{r 1bi_maltreat_stats}
deaths = c(11, 15, 15, 16, 25, 20)
m_deaths = mean(deaths)
s_deaths = sd(deaths)
var_deaths = var(deaths)
var_deaths_v2 = s_deaths^2
```
Over the period from 1990-1995, was a **mean** of `r m_deaths` juvenile deaths per year caused by maltreatment. The **variance** in these maltreatment deaths was `r var_deaths`. The **standard deviation** of the same maltreatment deaths was `r s_deaths`.

### 1.b-ii)
**Calculate standard error of the mean:**
```{r 1bii_maltreat_se}
n = length(deaths)
se_deaths = s_deaths/(sqrt(n))
```
For a sample of `r n`, the **standard error** of $latex {\bar x}$ = `r se_deaths`.

### 1.b-iii)
The algebraic equation for the estimated number (`T`) of maltreatment deaths in the period of `k` years is: 

T = `r m_deaths` * k

As `k` increases, the total (T) increases linearly as a function of the mean number of deaths (i.e., `r m_deaths`).

### 1.b-iv)
The standard deviation of the statistic (T) = `k *` sd(X). In this case, sd(X) is equal to the **standard error** of the sample mean ($latex {\bar x}$), calculated in **1b-ii**; thus sd(T) = `k *` `r se_deaths`. 

### Example distributions using linear transformation:
```{r example_create_distributions}
k = 1
year1 <- data.frame(length = rnorm(1000, mean = m_deaths*k, se_deaths*k))
k = 2
year2 <- data.frame(length = rnorm(1000, mean = m_deaths*k, se_deaths*k))
k = 3
year3 <- data.frame(length = rnorm(1000, mean = m_deaths*k, se_deaths*k))
```

```{r plot distributions, fig.width=9, fig.height=6}
#Now, combine your two dataframes into one.  First make a new column in each.
year1$NumberYears <- '1 year'
year2$NumberYears <- '2 years'
year3$NumberYears <- '3 years'

#and combine into your new data frame all_years
all_years <- rbind(year1, year2, year3)

library(ggplot2)
#now make your histogram
ggplot(all_years, aes(length, fill = NumberYears)) + geom_histogram(alpha = 0.5, colour = "darkgray", aes(y = ..density..), position = 'identity')
```

Question 2
-----------------------------------------

### 2.a)
Here we will use the data in `earlydeaths.csv` to generate a **cross-tabulation**. This will indicate if there is a relationship between the variables `time` and `cause` (cause of death), which we will formally test in **2b**.

**Load in data:**
```{r 2a_earlydeaths_load}
# load in data
d0 = read.csv('earlydeaths.csv')
str(d0)
summary(d0)
```

**Change levels of `cause` to match the cross-tab in homework:**
```{r 2a_earlydeaths_editlevel}
levels(d0$cause)
levels(d0$cause)[1] = 'Maltreat'
levels(d0$cause)[2] = 'Non-Maltr'
levels(d0$cause)
```

**Generate cross-tabulation of `time` by `cause`:**
```{r 2a_earlydeaths_crosstab}
# Print the table
print(table(d0$cause, d0$time))

# Print table with margins, like in hw
tx = addmargins(table(d0$cause, d0$time, dnn=c('Cause','Time')))
print(tx)
```


### 2.b)
```{r 2a_earlydeaths_chisq}
# Print the table summary (i.e., chi-squared contingency test)
rs1 = summary(table(d0$cause, d0$time))
print(rs1)

# access variables from this output "object"
chisq = rs1$statistic
df = rs1$parameter
pval = rs1$p.value
```

To test the null hypothesis that there is no relationship between the cause of juvenile deaths (`cause`) and time (`time`), we conducted a **chi-squared contingency test**. Results indicate that there is not a significant relationship between the cause of deaths and time, $latex {\chi^2}$(`r df`) = `r chisq`, *p* = `r pval`. In other words, the time trend in the number of deaths is similar regardless of the cause of death.

**n.b., you can write stats output in the following ways:**
$latex {\chi^2}$(`r df`) = `r chisq`, *p* = `r pval`;
X^2 (2) = 0.43, *p* = 0.81.

If you decide to use latex, [this site](http://www.calvin.edu/~rpruim/courses/m343/F12/RStudio/LatexExamples.html) is a helpful resource!

Question 8
-----------------------------------------
### 8.a)
**Create variable storing population size for age ranges:**
```{r 8a_createvar}
census_pop = c(72.3, 46.5, 43.2, 42.5, 41.9, 35)
```

**Calculate size of sub-population >18 years old:**
```{r 8a_18overpop}
pop_Over18 = sum(census_pop) - census_pop[1]
```
The size of the sub-population of likely voters is `r pop_Over18`.

**Calculate proportion of likely voters by age category:**
```{r 8a_p_likelyvoter_byage}
p_likelyvoters = census_pop[2:6]/pop_Over18
```
The proportions of likely voters for the age categories over 18 are: `r p_likelyvoters`.

### 8.b)
**Load in data from fieldsimul1.csv:**
```{r 8b_load_fieldsimul1}
d1 = read.csv('fieldsimul1.csv')
str(d1)
summary(d1)
```

**Recode age into categorical variable using 5 age categories:**
```{r 8b_agecat0}
d1$agecat0 = findInterval(d1$age, c(30,40,50,66))
d1$agecat0 = d1$agecat0 + 1     # group 1 starts at 1
d1$agecat0 = factor(d1$agecat0) # convert to categorical var
head(d1)
```

### 8.c)

```{r 8c_chisq}
rs4 = chisq.test(table(d1$agecat0), p = p_likelyvoters, simulate.p.value = FALSE)
print(rs4)

# Note this format also gives the same results:
rs4_v2 = chisq.test(table(d1$agecat0), p = census_pop[2:6], rescale.p = TRUE, simulate.p.value = FALSE)
print(rs4_v2)
```

Explain results....

```{r 8c_plot_proportions, fig.width = 7, fig.height=5}
p_observed = table(d1$agecat0)/length(d1$agecat0)
p_expected = p_likelyvoters

df_cat =rbind(p_observed, p_expected)
barplot(df_cat, col=c("darkblue","red"), xlab = ('Age Category'),
ylab = 'Proportion', legend = rownames(df_cat), beside=TRUE)
```

### 8.d)
```{r 8d_levels_recall_party}
d1$party = factor(d1$party)
levels(d1$party) = c('Dem', 'Rep', 'Other')

d1$recall = factor(d1$recall)
levels(d1$recall) = c('Yes', 'No', 'Unsure')

head(d1)
```

```{r 8d_crosstab_recall_party}
table(d1$recall, d1$party)

# Print table with margins
tx = addmargins(table(d1$recall, d1$party, dnn=c('Recall','Party')))
print(tx)

# Run chi-squared test
rs5 = chisq.test(table(d1$recall, d1$party))
print(rs5)
```

Explain results...

```{r 8d_crosstab_minus_unsure}
d3 = d1[d1$recall != 'Unsure',]
str(d3) # note the 3 levels of recall still!
d3$recall = factor(d3$recall)

rs6 = chisq.test(table(d3$recall, d3$party))
print(rs6)
```

Explain results...

## 8.e)
**Re-code recall as quantitative variable:**
```{r 8e_recall_quantitative}
d1$recallq = as.numeric(d1$recall)
d1$recallq = 1*as.numeric(d1$recallq==1) + (-1)*as.numeric(d1$recallq==2) + 0*as.numeric(d1$recallq==3)

summary(d1)
```

## 8.f)
```{r 8f_plot_recallq-vs-agecat0_byparty, fig.width=9, fig.height=7}
interaction.plot(d1$agecat0, d1$party, d1$recallq,
                 xlab = 'Age Category', ylab = 'Recall')
```

## 8.g)
```{r 8g_lm_recallq_by_agecat-and-party}
str(d1)

# additive model
rs_add = lm(recallq ~ agecat0 + party, data = d1)
summary(rs_add)

# interactive model
rs_inter = lm(recallq ~ agecat0 * party, data = d1)
summary(rs_inter)

# quadratic trend of age?
d1$agecat0 = as.numeric(d1$agecat0)
str(d1)

# additive model w/quadratic effect of age category
rs_add_polyage = lm(recallq ~ poly(agecat0, 2) + party, data = d1)
summary(rs_add_polyage)

# interactive model w/quadratic effect of age category
rs_inter_polyage = lm(recallq ~ poly(agecat0, 2) * party, data = d1)
summary(rs_inter_polyage)
```



Explain output...



small_sample = sample(D0, 7, replace=T, prob = p0)
muSS = mean(small_sample)
sdSS = sd(small_sample)
hist(small_sample)
print(muSS)
print(sdSS)
