Template Homework
========================================================

To submit your homework, you'll create an RMarkdown (.Rmd) script to upload to Coursework. This will allow the TAs to create an easy-to-read document to grade; more importantly, you will also get to practice *(1)* writing scripts, *(2)* keeping track of the analyses you run, and *(3)* organizing your output.

Below are some basics for using RMarkdown (also, check out [this website](http://www.rstudio.com/ide/docs/authoring/using_markdown) for more info!).

One thing that's nice about RMarkdown is that you can clearly differentiate **sections of code** from **commentary about what you are trying to do**. When you want to include sections of code, you mark them off by using the header and footer below.

Inserting an R Script:
```{r}
summary(cars)
```

You can name the "chunks" of code so that you remember what each section of code does. Then, you can jump to chunks of code (which makes it easy to search through your document), and run each chunk separately. Below is the same chunk of code as above with a "name" included:

```{r example chunk}
summary(cars)
```

You can also embed plots, for example:

```{r fig.width=7, fig.height=6}
plot(cars)
```

In this case, the information included in the **"name"** of the chunk specifies the dimensions of the plot.

After creating a .Rmd document, you can easily convert the script into HTML by hitting the **"Knit HTML"** button - try it out and see how nice it looks!

In the document `Template_HW.R` we have worked through Question 9 Parts A&B so you get an idea of how to format your homework.



