Plotting Data by Facets
========================================================

Sometimes you might want to split your data by one (or more) variables, and visualize the subsets of the data together. Facets allow you implement this using `ggplot`.

Load in data
-------------
```{r fig.width = 6, fig.height = 4}
data = ChickWeight
```

Draw basic scatter plot w/linear regression
----------------------
```{r fig.width = 6, fig.height = 4}
library(ggplot2)

# change background to white, increase font size
theme_set(theme_bw(base_size = 18)) 

# set up basic scatter plot, overlaid with linreg
pl = ggplot(data, aes(x=Time, y=weight)) +
  geom_point(shape=16, 
             color = 'black',
             alpha = 0.5) +
  geom_smooth(method=lm, color = 'red')
```

Facet by a variable
----------------------
```{r fig.width = 6, fig.height = 4}
pl + facet_grid(~Diet)
```

Plot a subset of the subjects by facet
----------------------

###Subset the data
Create a function to create a smaller subset of data with fewer subjects.

```{r}
# create a function that will subset the data
subsetData = function(data, sub_col, subset_num){
  # data: dataset
  # sub_col = column with subj ids
  # sample_num = num unique subj for subset
  
  # list of unique subjects
  unique_subs = unique(data[,sub_col])
  
  # take a subset of subjects' data
  subs_subset = sample(unique_subs, subset_num)
  
  # subset the data based on the subset
  data_subset = data[data[,sub_col] %in% subs_subset,]
  
  return(data_subset)
}

# number to subset
subset_num = 6

data_subset = subsetData(data, 'Chick', subset_num)
summary(data_subset)
```

Now plot the subset of subjects with `facet_wrap()`, coloring the regression line by Diet. Note, you can limit the number of subplots per row by specifying the number of columns with `ncol`.

```{r}
pl = ggplot(data_subset, aes(x=Time, y=weight, color=Diet)) +
  geom_point(shape=16, 
             alpha = 0.5) +
  geom_smooth(method=lm)

pl + facet_wrap(~Chick, ncol=3)
```
