Psych 252 Review
========================================================

Here we will analyze data from a study of 2287 eighth-grade pupils (aged about 11) in 132 classes in 131 schools in the Netherlands. 

The variables in the dataframe are:
- `lang`: language test score
- `IQ`: verbal IQ
- `class`: class ID
- `GS`: class size (number of eighth-grade pupils recorded in the class (there may be others see COMB, and some may have been omitted with missing values)
- `SES`: social-economic status of pupil's family
- `COMB`: were the pupils taught in a multi-grade class (0/1)? Classes which contained pupils from grades 7 and 8 are coded 1, but only eighth-graders were tested

Load in the data
----------------------
```{r}
library(MASS)
data(nlschools)
data(n) # use tab-complete to see details
df = data.frame(nlschools)
str(df)
# is everything factored correctly?
```

Take a quick look at relationships between some variables:
----------------------
```{r}
df_corr = data.frame(cbind(df$lang, df$IQ, df$SES, df$GS))
colnames(df_corr) = c("langScore", "IQ", "SES", "classSize")
z <- cor(df_corr)

require(lattice)
levelplot(z, panel = panel.levelplot.raster, par.settings = list(regions = list(col = heat.colors(100))))
```

Now, let's ask some questions, and examine them using our tools from stats!:
----------------------

### Does IQ predict language test score?
```{r}
library(lme4)
rs0 = lm(lang ~ scale(IQ, scale=FALSE), data = df)
summary(rs0)
rs1 = lmer(lang ~ scale(IQ, scale=FALSE) + (1|class), data = df)
summary(rs1)
rs2 = lmer(lang ~ scale(IQ, scale=FALSE) + (1 + IQ |class), data = df)

# test models..
anova(rs0, rs1)
anova(rs1, rs2)

# or manually!
dev1 = -2*as.numeric(logLik(rs0))
dev2 = -2*as.numeric(logLik(rs1))

# get chisq p-stat (since diff in deviance has chi-sq distribution)
diff = dev1-dev2; diff
4 - length(rs0$coefficients)
pchisq(diff, df=2, lower.tail=FALSE)

summary(rs2)
coef(rs2)
```

### Does IQ predict language test score, when controlling for SES?
```{r}
rs3 = lmer(lang ~ scale(IQ, scale=FALSE) + scale(SES, scale=FALSE) + (1 |class), data = df)
summary(rs3)
```

### Does combined class predict language test score?
```{r}
df$COMB_fac = factor(df$COMB)
levels(df$COMB_fac)

rs4 = lmer(lang ~  COMB_fac + (1 |class), data = df)
summary(rs4)

contrasts(df$COMB_fac) = c(1, -1)
#contrasts(df$COMB_fac) = c(-1, 1)
#contrasts(df$COMB_fac) = c(0, 1)
contrasts(df$COMB_fac)
rs4_v2 = lmer(lang ~  COMB_fac + (1 |class), data = df)
summary(rs4)

plot(lang ~  COMB_fac, data=df)
aggregate(lang ~ COMB_fac, data=df, mean)
```

### Are there similar numbers of students observed from each class?
```{r}
tb = table(df$class)
tb
chisq.test(tb)
```

### Were there similar numbers of students from combined/not combined classes?
```{r}
tb = table(df$COMB)
tb
chisq.test(tb)
prop.table(tb)
```

### Does SES predict whether or not a student is in a combined class?
```{r}
rs_glm = glmer(COMB ~ SES + (1 | class), family='binomial', data = df)
rs2_glm = glmer(COMB ~ SES + (1 + SES | class), family='binomial', data = df)
anova(rs_glm, rs2_glm)
summary(rs2_glm)
```